---
title: 'Report: Analysis of NOAA weather data for severity of population and economic
  impact'
output:
  pdf_document: default
  html_document:
    keep_md: yes
'knitr :': opts_chunk$set (warning=FALSE)
---

## Introduction
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This report involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The basic goal of this report is to explore the NOAA Storm Database and answer some basic questions about severe weather events as below:

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

The analysis in this report summarizes the top 10 weather events causing the highest impact to population and economy factors such as property and crops. The analysis can provide indicators on how best to prioritize resources for relief.      

  
### 0. Install packages


```{r}
packages <- c('ggplot2','dplyr','zoo','reshape2')
check_packages <- function (...) {
  lapply(list(...), function (p) {
     if (! p %in% installed.packages()) {
        install.packages(p)
        require(p,character.only=TRUE)
     }
  })
}

```


### 1. Download storm data & convert column names to lower case for consistency


```{r}

zipfilename <- "repdata%2Fdata%2FStormData.csv.bz2"

fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"


if (!file.exists(zipfilename)) {
    download.file(fileURL, dest=zipfilename,mode="wb",quiet=FALSE) 
   }

df <- read.csv(bzfile("repdata%2Fdata%2FStormData.csv.bz2"),header=TRUE,stringsAsFactors = FALSE)

colnames(df) <- tolower(colnames(df)) 

```


### 2. Data Processing & transformation

The following steps prepare the data for processing. The conversion from dmgexp (codes) to the multiplication factor is needed to compute costs. For example the value 'k' corresponds to 1000's.

* prepare a lookup table to convert prop damage codes  
  ("" "-" "?" "+" "0" "1" "2" "3" "4" "5" "6" "7" "8" "B" "h" "H" "K" "m" "M") 
  to multiplication factors  
  (0,0,0,1,10,10,10,10,10,10,10,10,10,109,102,102,103,106,106)
* The above multplication factors are applied to derive property and crop damage costs



```{r}

df.damage <- subset(df,select=c(evtype, propdmg,propdmgexp,cropdmg,cropdmgexp))
Symbol <- sort(unique(as.character(df.damage$propdmgexp)))
Factor <- c(0,0,0,1,10,10,10,10,10,10,10,10,10,10^9,10^2,10^2,10^3,10^6,10^6)

convert.Factor <- data.frame(Symbol, Factor)

df.damage$Prop.Factor <- convert.Factor$Factor[match(df.damage$propdmgexp, convert.Factor$Symbol)]
df.damage$Crop.Factor <- convert.Factor$Factor[match(df.damage$cropdmgexp, convert.Factor$Symbol)]

```


### 3. Analysis question 1 - Most harmful events to population health

The code below processes the data to determine which type of events are most harmful with respect to population health. 

The results show the top 10 event types causing fatalities and injuries


```{r}

library(dplyr)

hdf <- aggregate(data=df,cbind(fatalities,injuries) ~ evtype,FUN=sum) %>% arrange (desc(fatalities))
head(hdf,10)

```

The results show that Tornado's are by far the biggest cause of fatalities and injuries followed by Excessive heat and Flash floods.


### 4. Analysis question 2 - Events causing greatest economic consequences


The following steps are executed:

* Compute property and  crop damage costs
* Compute total damage cost as of property and crop damagae costs
* The top 10 costs by total damage is produced as output

```{r}

df.damage <- df.damage %>% mutate(propdmg = propdmg*Prop.Factor, cropdmg = cropdmg*Crop.Factor, total.dmg = propdmg+cropdmg)

edf <- aggregate(data=df.damage,cbind(total.dmg,propdmg,cropdmg) ~ evtype,FUN=sum) %>% arrange (desc(total.dmg))
 
head(edf,10)

```

The results show that Floods are the biggest cause of total damage arising out of property and crop damage costs


### 5. Figure - Analysis on damage impact of top 10 weather events to population
```{r}
library(ggplot2)
library(reshape2)
par(mfrow=c(1,1))
dfmelt <- melt(hdf[1:10,],id.vars='evtype')
ggplot(dfmelt,aes(x=reorder(evtype,-value),y=value)) + geom_bar(stat='identity',aes(fill=variable),position='dodge') +  theme(axis.text.x = element_text(angle=45,hjust=1))+ggtitle("Top 10 Events with highest population impact") +labs(x="EVENT TYPE", y="Total impact")


```



### 6. Figure - Analysis on damage impact of top 10 weather events to economy

```{r}
library(reshape2)
edfmelt <- melt(edf[1:10,],id.vars='evtype')
ggplot(edfmelt,aes(x=reorder(evtype,-value),y=value)) + geom_bar(stat='identity',aes(fill=variable),position='dodge') +  theme(axis.text.x = element_text(angle=45,hjust=1))+ggtitle("Top 10 Events with Highest impact") +labs(x="EVENT TYPE", y="Total economic impact")

```

### 7. Summary of Results

This report analysis the weather data over many years to produce a report on a) which events are the most impactful to the population b) which events produce the most impacful economic consequences.

The report details the top events for both these analysis including tables and charts showing quantitatively and visually the top events

* the analysis determines that tornadoes are the largest cause of fatalities and injuries
* the analysis determines that Floods are largest cuase of economic consequences


